---
title: "Honours Project"
author: "Ben van Huyssteen & Dylan Rowell"
date: "2023-04-01"
output: 
  pdf_document: default
  html_document: default
header-includes: \usepackage{pdfpages} \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```


# Brief and background (Colin meeting notes)

This project deals with estimating the (instant) mortality rate of 10
different Southern African fish species based on the length
distribution.

Species include: Galjoen, Blacktail, Dusky Cob, Elf, Zebra, White
Steenbras, Spotted Gully shark, Belman, Guitarfish, White Stumpnose.

The mortality rate of fish is defined as ...

The mortality rate (M) can be derived from the age distribution of a
species. The age distribution quantifies how many fish are alive in each
age bracket. The age of a fish can be determined by catching and
dissecting the fish and examining biological age indicators. This is not
a sustainable and ethical \*better words\* way of estimating mortality
rates, especially for rare and vulnerable species.

Fish length and age are linked, but naturally do not correspond exactly;
there is variation. For example, two 3-year-old fish of a species will
probably have different lengths, and two 30cm fish could be different
ages based on how fast they grow.

We are given length-frequency data from the De Hoop MPA, and aim to use
this to estimate the natural mortality rate. As we have seen, this
length distribution cannot be used directly to measure M, but needs to
be used to estimate the age distribution first, from which we can then
estimate M. This length-age relationship has been historically measured
and represented with age-length keys for each species. Via von
Bertalanffy models, this relationship has been modelled.

## Von Bertalanffy models

\*\*not researched, just from meeting For each fish age, there is a
normal distribution of lengths. As age increases, the variance of the
length estimate increases too due to more natural variation as well as
there being less data at those ages.

```{r blacktail EDA}
blacktail = dat[dat$SpeciesName == "Diplodus sargus",] 
blacktail

#von Bertalanffy 

L_inf = 403
k = 0.142
t_0 = -1.69

VBL_length = function(age, L_inf,k,t_0){
  return(L_inf*(1-exp(-k*(age-t_0))))
}

VBL_age = function(len, L_inf,k,t_0){
  return(-log(1-len/L_inf)/k+t_0)
}

ages = seq(0,30, length = 50)
plot(VBL_length(ages, L_inf,k,t_0)~ages, type = "l", col = "blue")+grid()

lengths = seq(0,402,length = 50)
plot(VBL_age(lengths, L_inf,k,t_0)~lengths, type = "l", col = "red")+grid()


```

## Possible methods (Res meeting notes)

-bootstrapping: repeatedly predict the age from the length data & fit
the age-length curve multiple times, each time getting M. Will get
confidence intervals from the bootstrapping

-hierarchichal bayes: original length-age & new data, can account for
uncertainty (??)

-use raw data from age-length key if available: to fit model ourselves

## Proposal Guidlines

detailing research topic itself and the analysis to be conducted, division of labour, and a time-line scheduling goals/tasks/deliverables to be completed 

- Describe problem, approach , 
- describe data, summary stats
  - include info on fish species &how data was collected
- possible methods. 
- Timeline (plan) with milestones, time frame, changes, anything else we might want to take into account
- application of project






# Proposal

## Outline

The natural mortality rate, M, for a species of fish is the instantaneous rate of death due to all factors except fishing, such as predation, disease and starvation (Mannini A et.el elephant in room). 
A value for M can be derived from the age distribution of a species, which quantifies how many fish are alive in each age bracket. This age distribution can only be directly determined by aging a large number of fish by examining biological age indicators, a cost and resource-intensive process which requires the dissection and thus death of the animal. (reference)

Especially for rare or vulnerable fish species, this is not feasible; a method that doesn't involve directly aging the fish to estimate M is required. We will explore methods of estimating M using the *length* distribution of a species. This project deals with estimating the (instant) mortality rate M with error of 10 different Southern African fish species based on their length distributions.

## Data

*something on how data was collected

We are supplied with data on many different fish species from the De Hoop Marine Protected Area (MPA) from 1948 to 2021. This is represented in an excel spreadsheet with the following column names: Year, Location, SpeciesName, Length. This is a convenient representation as the data is already in long format, and allows us to readily extract the lengths for each species. The location variable is one of "Koppie Alleen" or "Lekkerwater", and is not of much importance; all data was collected in the same reserve (maybe put some EDA graphs here to show this. lengths of fish in each location). The Year, SpeciesName and Length are self-explanetory, giving the year the fish was caught and measured, scientific name and length in cm respectively.

*selectivity in data making it look normal but actually exponential

```{r data}
rm(list = ls())
setwd("C:/Users/benva/Desktop/HON STATS/PROJECT/Stats-Honours-Project")

library("readxl")

dat = read_xlsx("qryLengthofDeHoopFishSpecies.xlsx", sheet = 2, col_names = c("Year","Location","SpeciesName","Length"),range = cell_limits(c(2, 1), c(NA, 4)))

colnames(dat)

min(dat$Year)
max(dat$Year)

plot(as.factor(dat$Location))

```

```{r cob}

cob = dat[dat$SpeciesName == "Argyrosomus japonicus",] 
cob

hist(cob$Length)

```




## Methods

The age of a fish is correlated to it's length (the length between the tip of its head and the fork in its tail**), but naturally do not correspond exactly; there is variation in the population based on how fast they grow. For example, two 3-year-old fish of a species will likely have different lengths, and two 30cm fish could be different ages. Using the length distribution to estimate the age distribution is the first task and actually the crux of the problem. Once the age distribution has been estimated, it is just a case of fitting an exponential regression model to the data to estimate M (see the section on exponential regression).  

### Von Bertalanffy Growth Model

The Von Bertalanffy Growth Model is the most commonly used model to model fish length-at-age. It is given as
$$
L(a) = L_\infty (1-e^{-k(a-t_0)})
$$
where $L(a)$ is the length of the fish at age a, $L_\infty$ is the average maximum length or asymptotic size, $k$ is the growth coefficient, and $t_0$ is the theoretical length at age 0. 

*parameters estimated using Levenberg-Marquardt Method for non-linear regression (http://www.pisces-conservation.com/growthhelp/index.html?von_bertalanffy.htm)

*historical data from age-length keys used


### Age-Length Keys from Historical Data

### Exponential Regression



# notes

-having issues reading in massive excell file - fixed by googling -
apparantly a bug in "cellranger", workaround found. -white see bream
Diplodus sargus has lots of data, maybe do that one?
-try to fit VBL model using tranformed linear regreassion: rewrite model as linear with logs and shit, then fit to original length-age data: will then get errors