---
title: "Honours Project Proposal"
author: "Ben van Huyssteen & Dylan Rowell"
date: "2023-04-01"
output: 
  pdf_document: default
  html_document: default
header-includes: \usepackage{pdfpages} \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

# Proposal Guidlines

detailing research topic itself and the analysis to be conducted, division of labour, and a time-line scheduling goals/tasks/deliverables to be completed 

- Describe problem, approach , 
- describe data, summary stats
  - include info on fish species &how data was collected
- possible methods. 
- Timeline (plan) with milestones, time frame, changes, anything else we might want to take into account
- application of project

# Introduction

The natural mortality rate, M, for a species of fish is the instantaneous rate of death due to all factors except fishing, such as predation, disease and starvation (Mannini A et.el elephant in room). 
A value for M can be derived from the age distribution of a species, which quantifies how many fish are alive in each age bracket. This age distribution can only be directly determined by aging a large number of fish by examining biological age indicators, a cost and resource-intensive process which requires the dissection and thus death of the animal. (reference)

Especially for rare or vulnerable fish species, this is not feasible; a method that doesn't involve directly aging the fish to estimate M is required. We will explore methods of estimating M using the *length* distribution of a species. This project deals with estimating the (instant) mortality rate M with error of 10 different Southern African fish species based on their length distributions.

# Data

We are supplied with data on many different fish species from the De Hoop Marine Protected Area (MPA) from 1948 to 2021. This is a 289 km2 MPA spanning 48km of coastline between Witsand and Skipskop that extends 5.6 km out to sea.

The data is represented in an excel spreadsheet with the following column names: Year, Location, SpeciesName, Length. This is a convenient representation as the data is already in long format, and allows us to readily extract the lengths for each species. 

The location variable is one of "Koppie Alleen" or "Lekkerwater", and is not of much importance; all data was collected in the same reserve. The box-plots for length at each location for cob and blacktail are shown to demonstrate how the location is negligible. The Year, SpeciesName and Length are self-explanetory, giving the year the fish was caught and measured, scientific name and length in mm respectively.

```{r data}
rm(list = ls())
setwd("C:/Users/benva/Desktop/HON STATS/PROJECT/Stats-Honours-Project")

library("readxl")

dat = read_xlsx("qryLengthofDeHoopFishSpecies.xlsx", sheet = 2, col_names = c("year","location","species","length"),range = cell_limits(c(2, 1), c(NA, 4)))

#colnames(dat)

#min(dat$Year)
#max(dat$Year)

cob = na.omit(dat[dat$species == "Argyrosomus japonicus",])

blacktail = na.omit(dat[dat$species == "Diplodus sargus",])
blacktail = blacktail[-which.max(blacktail$length),] #removing magnum blacktail
max(blacktail$length)

#numbers at location overall
plot(as.factor(dat$location))

#boxplot cob & blacktail at location
par(mfrow = c(1,2))
plot(cob$length~as.factor(cob$location), xlab = "Location", ylab = "Cob Length")
plot(blacktail$length~as.factor(blacktail$location), xlab = "Location", ylab = "Blacktail Length")

```

## Collection Method

The data has been collected by scientists with permits to catch and measure fish in the MPA since 1981. Teams catch fish with a fishing rod, record its where it was caught, its species and its tail length. The length measurement is taken by measuring the distance between the tip of the fishes mouth to the fork in its tail. **measurement error

## 

*selectivity in data making it look normal but actually exponential

## Species considered

```{r species table}


tab = read_xlsx("Species_table.xlsx")
tab$
knitr::kable(tab)
```


```{r cob}

cob = dat[dat$SpeciesName == "Argyrosomus japonicus",] 
cob

hist(cob$Length)

```


# Methods

The age of a fish is correlated to it's length (the length between the tip of its head and the fork in its tail**), but naturally do not correspond exactly; there is variation in the population based on how fast they grow. For example, two 3-year-old fish of a species will likely have different lengths, and two 30cm fish could be different ages. Using the length distribution to estimate the age distribution is the first task and actually the crux of the problem. Once the age distribution has been estimated, it is just a case of fitting an exponential regression model to the data to estimate M (see the section on exponential regression).  


## Age-Length Keys from Historical Data

*historical data from age-length keys used

## Von Bertalanffy Growth Model

The Von Bertalanffy Growth Model is the most commonly used model to model fish length-at-age. It is given as
$$
L(a) = L_\infty (1-e^{-k(a-t_0)})
$$
where $L(a)$ is the length of the fish at age a, $L_\infty$ is the average maximum length or asymptotic size, $k$ is the growth coefficient, and $t_0$ is the theoretical length at age 0. 

*parameters estimated using Levenberg-Marquardt Method for non-linear regression (http://www.pisces-conservation.com/growthhelp/index.html?von_bertalanffy.htm)

## Exponential Regression

*used to get m from age dist


# 

